Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2018-05-14 conclusions.tex
Minor wording changes

As introduced above, hack weeks serve multiple purposes, including dissemination of state-of-the-art technological advances through the scientific community, building collaborations between academic subdisciplines and fostering interdisciplinary research as well as  promoting open science and reproducibility.
Initial results from three events held in 2016 and 2017 in three different fields (astronomy, geosciences and neurosciences) indicate that hack weeks succeed at all of these objectives, but that the measure of success is field-specific in that it depends to some degree on how much the concepts hack weeks promote were already adopted within the community.
Hack weeks are still a very young concept, and estimating the long-term impact of these events within the scientific communities they serve will require follow-up over multiple years to asses their effect on collaboration networks, career outcomes for early-career academics and adoption of new methods.
We have shown, however, that hack weeks provide an easy-to-implement, fairly low-cost method to introduce new technologies and methods into scientific fields on much shorter time scales than traditional teaching efforts can.
While we focus here on hack weeks in scientific fields, the concept could be extended to other areas, and is more generally useful in any area (1) where useful tools can be learned in short tutorials, (2) where results and outcomes can be produced on the timescale of a few days, and (3) that would benefit from collaborative approaches that cross traditional boundaries. Such areas could include for example the social sciences, music and art.

>>>>

As introduced above, hack weeks serve multiple purposes, including dissemination of state-of-the-art technological advances through the scientific community, building collaborations between academic subdisciplines and fostering interdisciplinary research as well as  promoting open science and reproducibility.
Initial results from three events held in 2016 and 2017 in three different fields (astronomy, geosciences and neurosciences) indicate that hack weeks succeed at all of these objectives.

Hack weeks are still a very young concept, and estimating the long-term impact of these events within the scientific communities they serve will require follow-up over multiple years to asses their effect on collaboration networks, career outcomes for early-career academics and adoption of new methods.
We have shown, however, that hack weeks provide an easy-to-implement, fairly low-cost method to introduce new technologies and methods into scientific fields on much shorter time scales than traditional teaching efforts can.
While we focus here on hack weeks in scientific fields, the concept could be extended to other areas, and is more generally useful in any area (1) where useful tools can be learned in short tutorials, (2) where results and outcomes can be produced on the timescale of a few days, and (3) that would benefit from collaborative approaches that cross traditional boundaries. Such areas could include for example the social sciences, music and art.

### 2018-05-14 intro.tex
Minor wording changes

%\section*{}
\label{sec:introduction}
\dropcap{A}s data become cheaper to gather and store, researchers have become increasingly reliant on computational workflows requiring skills in statistical modeling, machine learning and scalable computation. In addition, the recent reproducibility crises in several scientific fields has motivated the acquisition of skills in open science and the design of reproducible workflows \cite[e.g.][]{pashler2012,baker2016}.
Formal university curricula have been relatively slow to offer courses in these important topics: the slack in this area has often been picked-up by extra-curricular, ad-hoc efforts such as workshops \cite{demasi2017}.
Well-known examples are the Software and Data Carpentry workshops providing training in research computing skills through a volunteer instructor program \cite{b:wilson-swc-lessons-2016,teal2015data}.
At the same time, there has been a rise in the number of statistical and computational courses designed for specific scientific disciplines.
Examples include the \textit{Summer School in Statistics for Astronomers}\footnote{\url{http://astrostatistics.psu.edu/su16/}}, the Google Earth Engine User Summits\footnote{\url{https://events.withgoogle.com/google-earth-engine-user-summit-2017/}}, as well as a variety of project-focused (rather than pedagogical) meetings, such as the dotAstronomy meetings\footnote{\url{http://dotastronomy.com}}.
Shorter, but similar-spirit meetings have been held in conjunction with conferences, such as the Hack Days at the annual American Astronomical Society meetings, the Brainhack hackathons associated with the Organization for Human Brain Mapping and the Society for Neuroscience\cite{Cameron_Craddock2016-wc}, and a hackathon at the American Geophysical Union meeting\footnote{\url{http://onlinelibrary.wiley.com/doi/10.1002/2014EO480004/pdf}}. 
In general, many of these events tend to emphasize either a more traditional pedagogical class and lecture methodology, or emphasize a focus on ad-hoc projects that participants work on during the event (Figure \ref{fig:hackspectrum}).
Pedagogically-focused events (e.g., Software Carpentry workshops \cite{b:wilson-swc-lessons-2016} in which no student-led projects are pursued) follow a classic academic model where novices learn new skills from experts. This model tends to focus on a one-way flow of information from instructor to student, and are usually targeted towards participants in the training phase of their career. On the other end of the spectrum, project-focused workshops emphasize collaborative activities using existing skills, leading to the common perception that they are designed for technical experts, and this may limit their audience.
To bridge this gap, we describe here a model that we have implemented: ``Hack Weeks'' that aim to capitalize on the advantages of each of these models.
These week-long events combine structured periods focused on pedagogy (often with an emphasis on statistical and computational techniques) with less structured periods devoted to hacks and creative projects, with the goal of encouraging collaboration and learning among people at various stages of their career.

>>>>

%\section*{}
\label{sec:introduction}
\dropcap{A}s data become cheaper to gather and store, researchers have become increasingly reliant on computational workflows requiring skills in statistical modeling, machine learning and scalable computation. In addition, the recent reproducibility crises in several scientific fields has motivated the acquisition of skills in open science and the design of reproducible workflows \cite[e.g.][]{pashler2012,baker2016}.
Formal university curricula have been relatively slow to offer courses in these important topics: the slack in this area has often been picked-up by extra-curricular, ad-hoc efforts such as workshops \cite{demasi2017}.
Well-known examples are the Software and Data Carpentry workshops providing training in research computing skills through a volunteer instructor program \cite{b:wilson-swc-lessons-2016,teal2015data}.
At the same time, there has been a rise in the number of statistical and computational courses designed for specific scientific disciplines.
Examples include the \textit{Summer School in Statistics for Astronomers}\footnote{\url{http://astrostatistics.psu.edu/su16/}}, the Google Earth Engine User Summits\footnote{\url{https://events.withgoogle.com/google-earth-engine-user-summit-2017/}}, as well as a variety of project-focused (rather than pedagogical) meetings, such as the dotAstronomy meetings\footnote{\url{http://dotastronomy.com}}.
Shorter, but similar-spirit meetings have been held in conjunction with conferences, such as the Hack Days at the annual American Astronomical Society meetings, the Brainhack hackathons associated with the Organization for Human Brain Mapping and the Society for Neuroscience\cite{Cameron_Craddock2016-wc}, and a hackathon at the American Geophysical Union meeting\footnote{\url{http://onlinelibrary.wiley.com/doi/10.1002/2014EO480004/pdf}}. 
In general, many of these events tend to emphasize either a more traditional pedagogical class and lecture methodology, or emphasize a focus on ad-hoc projects that participants work on during the event (Figure \ref{fig:hackspectrum}).
Pedagogically-focused events follow a classic academic model where novices learn new skills from experts. This model tends to focus on a one-way flow of information from instructor to student, and are usually targeted towards participants in the training phase of their career. On the other end of the spectrum, project-focused workshops emphasize collaborative activities using existing skills, leading to the common perception that they are designed for technical experts, and this may limit their audience. 
To bridge this gap, we describe here a model that we have implemented: ``Hack Weeks'' that aim to capitalize on the advantages of each of these models.
These week-long events combine structured periods focused on pedagogy (often with an emphasis on statistical and computational techniques) with less structured periods devoted to hacks and creative projects, with the goal of encouraging collaboration and learning among people at various stages of their career. 
Our hack week model shares some similarities to those summer schools (e.g., Advanced Course on Computational Neuroscience, Okinawa Computational Neuroscience Course, Woods Hole Computational Neuroscience Summer course) that combine individual or group project work with teaching methods that transfer knowledge from experts to early-career researchers.
However, a distinguishing factor of our hack weeks in that they tend to be less structured and more participant-driven, so that the learning content is co-created between organizers and participants based on the unique experience of people in the room. 
This looser structure is designed to facilitate a multi-directional transfer of knowledge between a diverse group of people in different stages of their career and with different backgrounds (see also section Participants below). The flexibility of our approach results in shifts along the continuum of pedagogy and open project work across hack week disciplines, and from year to year within a specific discipline (Figure \ref{fig:hackspectrum}).

### 2018-05-14 participants.tex
Minor wording changes

For hack weeks, it is not obvious a priori that prerequisites for attendance in terms of skills should exist, and prerequisites will depend on the objectives of the workshop. For example, Astro Hack Week has traditionally accepted participants at all skill levels with respect to data science and did not include a merit-based selection, whereas Neuro Hack Week did include skill-based criteria in their selection (see also SM, Section 2.1 for more detail on the individual selection procedures).

If merit-based selection is part of the evaluation process, organizers face the crucial decision whether to assess merit blindly or not. Because human decision makers tend to be swayed by unrelated characteristics including name~\cite{bertrand2004} or gender~\cite{mossracusin2012}, an initial merit selection blinded to demographic characteristics can be an effective way to counteract certain biases related to internal heuristics. A merit selection could then be performed via scores given independently by members of the organizing committee based on a set of pre-defined, explicit selection criteria. This type of blinded procedure tends to reduce biases during participant selection when committees would otherwise not consider diversity during their selection [REF].

>>>>

For hack weeks, prerequisites will depend on the objectives of the workshop and may not exist at all. For example, Astro Hack Week has traditionally accepted participants at all skill levels with respect to data science and did not include a merit-based selection, whereas Neuro Hack Week did include skill-based criteria in their selection (see also SM, Section 2.1 for more detail on the individual selection procedures).

If merit-based selection is part of the evaluation process, organizers face the crucial decision whether to assess merit blindly or not. Because human decision makers tend to be swayed by unrelated characteristics including name~\cite{bertrand2004} or gender~\cite{mossracusin2012}, an initial merit selection blinded to demographic characteristics can be an effective way to counteract certain biases related to internal heuristics. A merit selection could then be performed via scores given independently by members of the organizing committee based on a set of pre-defined, explicit selection criteria. This type of blinded procedure tends to reduce biases during participant selection when committees would otherwise not consider diversity during their selection~\cite{bohnet2016}.

### 2018-05-14 results.tex
Minor wording changes

\begin{figure*}[h!]
%\begin{center}
%\begin{subfigure}[t]{\textwidth}
\centering
%\caption{}
\includegraphics[width=\textwidth]{f2.eps}
%\end{subfigure}
\caption{Post-workshop surveys from three hack weeks: participants in the 2016 astro-, geo- and neuro- hack weeks responded to questions assessing their experiences. The overall response rates for GHW and NHW were 100\%, with $N_{\mathrm{GHW}}= 83$ and $N_{\mathrm{NHW}} = 86$, respectively. At AHW, the combined overall response rate for both years was 76\%, or $N_{\mathrm{AHW}} = 72$ out of $94$ total attendees. Note that individual response rates may be slightly lower (as some questions were optional) and are reported in the panel titles for AHW, GHW and NHW, respectively. We report here about results in three different domains: the development of technical skills (a -- c), collaboration and teaching (d -- f), and shifts in attitudes towards reproducibility and open science (g, h).}
\label{fig:survey}
%\end{center}
\end{figure*}

@@

One important question regarding hack week outcomes is whether participants from underrepresented groups thrive at these events, or whether their full participation is impeded. Significant differences between minorities and non-minorities, even on a self-reported scale, for questions related to learning outcomes, teaching or network building would indicate that improvements in workshop facilitation and structure may be required to allow participants of all stripes to participate fully.
In Section 4 of the Supplementary Materials, we lay out various strategies we have employed to facilitate participation, empower members of underrepresented groups and mitigate impostor syndrome. Based on our surveys, we find no significant dependence of self-reported learning outcomes on gender identity or race/ethnicity (Table \ref{tab:survey}). Similarly, participants who self-identified as minorities with respect to gender identity and race or ethnicity do not significantly differ in their answers with respect to teaching outcomes for any of the hack weeks. Both underrepresented groups similarly feel that they built equally valuable connections at all three hack weeks and generally feel that their contributions to their hack teams were valued. For GHW, there is a weak indication that participants from racial/ethnic minorities respond more positively when asked about building connections (p=0.04) with a medium to large effect size ($\phi_c = 0.4$), while for AHW, responses regarding the value of contributions to hack teams may be spread more widely for participants from racial/ethnic minorities than for Caucasian participants ($p = 0.01; \phi_c = 0.4$). These initial results indicate that future work on how demographics interact with hack week attendance may be fruitful. We note that since we did not include a question about impostor syndrome itself, we cannot draw causal connections between our results and its prevalence. Additionally, our number of respondents is small, and the lack of significant differences may be caused by the small statistical power of our sample.
%It is notable, however, that we find no correlation between the response to this question and career stage for any of the events surveyed here. Participants at all career stages report similar engagement in teaching. This suggests there is some evidence for our hypothesis that compared to e.g.\ a traditional summer school, a hack week is less hierarchical and encourages lateral knowledge transfer between participants at different career stages. However, we note that the sample sizes per category are relatively small, and future work is needed to confirm this hypothesis. We similarly find no correlations with gender identity or race/ethnicity.

>>>>

\begin{figure*}[h!]
%\begin{center}
%\begin{subfigure}[t]{\textwidth}
\centering
%\caption{}
\includegraphics[width=\textwidth]{f2.eps}
%\end{subfigure}
\caption{Post-workshop surveys from three hack weeks: participants in the 2016 astro-, geo- and neuro- hack weeks responded to questions assessing their experiences. Response rates are reported in the panel titles for AHW, GHW and NHW, respectively. We report here about results in three different domains: the development of technical skills (a -- c), collaboration and teaching (d -- f), and shifts in attitudes towards reproducibility and open science (g, h).}
\label{fig:survey}
%\end{center}
\end{figure*}

@@

One important question regarding hack week outcomes is whether participants from underrepresented groups thrive at these events, or whether their full participation is impeded. Significant differences between minorities and non-minorities, even on a self-reported scale, for questions related to learning outcomes, teaching or network building would indicate that improvements in workshop facilitation and structure may be required to allow participants of all stripes to participate fully.
In Section 4 of the Supplementary Materials, we lay out various strategies we have employed to facilitate participation, empower members of underrepresented groups and mitigate impostor syndrome. Based on our surveys, we find no significant dependence of self-reported learning outcomes on gender identity or race/ethnicity (Table \ref{tab:survey}). Similarly, minority participants with respect to gender identity and race or ethnicity do not significantly differ in their answers with respect to teaching outcomes for any of the hack weeks. Both underrepresented groups similarly feel that they built equally valuable connections at all three hack weeks and generally feel that their contributions to their hack teams were valued. For GHW, there is a weak indication that participants from racial/ethnic minorities respond more positively when asked about building connections (p=0.04) with a medium to large effect size ($\phi_c = 0.4$), while for AHW, responses regarding the value of contributions to hack teams may be spread more widely for participants from racial/ethnic minorities than for Caucasian participants ($p = 0.01; \phi_c = 0.4$). These initial results indicate that future work on how demographics interact with hack week attendance may be fruitful. We note that since we did not include a question about impostor syndrome itself, we cannot draw causal connections between our results and its prevalence. Additionally, our number of respondents is small, and the lack of significant differences may be caused by the small statistical power of our sample.
%It is notable, however, that we find no correlation between the response to this question and career stage for any of the events surveyed here. Participants at all career stages report similar engagement in teaching. This suggests there is some evidence for our hypothesis that compared to e.g.\ a traditional summer school, a hack week is less hierarchical and encourages lateral knowledge transfer between participants at different career stages. However, we note that the sample sizes per category are relatively small, and future work is needed to confirm this hypothesis. We similarly find no correlations with gender identity or race/ethnicity.

### 2018-05-14 supplementary_material.tex
Minor wording changes

Here we explore a number of survey responses with gender identity as an independent variable.

@@

Here we explore a number of survey responses with racial/ethnic identity as an independent variable.

>>>>

Here we explore a number of survey responses with gender identity as an independent variable. We grouped ``gender identity'' into two categories: ``non-minority'' contains all participants who self-identified as male, ``minority'' contains all participants who responded with ``female'' or any other response. Women are here considered a minority with respect to their chosen field of study (neuroscience, astrophysics and geoscience), where they tend to be strongly underrepresented.

@@

Here we explore a number of survey responses with racial/ethnic identity as an independent variable. We grouped responses into two final categories, ``white'' (Caucasian participants) ``non-white'' (all others).

### 2018-05-14 themes.tex
Minor wording changes

To date, all hack weeks we have organized have been subject-specific, i.e.\ aimed at bringing together a community with a shared scientific interest, such as neuroscience.
Advantages to this approach include shared language and scientific objectives within communities organized by subject, leaving more time for active collaboration on cutting-edge science.
On the other hand, homogeneity may lead to \textit{group think} and inhibit new, creative solutions. 
In this case, it may be advantageous to design a hack week around a technique (e.g.\ Gaussian Processes) or modality (e.g.\ imaging), such as the ImageXD (image processing across domains\footnote{\url{http://http://www.imagexd.org/}} meetings. 
For these events, building a shared vocabulary and understanding of major data analysis problems is crucial, but they also allow for cross-disciplinary diffusion of techniques into other subjects and therefore decrease the risk of duplication of method development efforts.

>>>>

To date, all hack weeks we have organized have been subject-specific, i.e.\ aimed at bringing together a community with a shared scientific interest, such as neuroscience.
Advantages to this approach include shared language and scientific objectives within communities organized by subject, leaving more time for active collaboration on cutting-edge science.
On the other hand, homogeneity may lead to \textit{group think} and inhibit new, creative solutions. 
In this case, it may be advantageous to design a hack week around a technique (e.g.\ Gaussian Processes) or modality (e.g.\ imaging), such as the ImageXD (image processing across domains\footnote{\url{http://http://www.imagexd.org/}}) meetings. 
For these events, building a shared vocabulary and understanding of major data analysis problems is crucial, but they also allow for cross-disciplinary diffusion of techniques into other subjects and therefore decrease the risk of duplication of method development efforts.

### 2018-05-14 what.tex
Minor wording changes

Our hack weeks combine structured, tutorial-style instruction with informal education and peer learning opportunities occurring within projects and hacks. 
Our focus on both pedagogy and open project work aims to fill an educational space we feel is not currently addressed by existing models (Figure \ref{fig:hackspectrum}). 
The hackathon, a time-bounded, collaborative events that bring together participants around a shared challenge or learning objective \cite{Decker2015}, forms one primary axis of our events. 
Hackathons originated from the open-source software movement and have historically focused on software and technology development. 
In recent years hackathons have evolved into a model providing opportunities for intensive, interdisciplinary collaboration \cite{Groen2015-cj} and education \cite{Kienzler2015-zu,Lamers2014-xf} in the sciences. 
Core element of all hackathons include opportunities for networking, strengthening of social ties, and the building of community connections, both within and across disciplines.
Building on these core elements hackathons have been implemented in different ways depending on the overall purpose, mode of participation, style of work environment and participant motivation \cite{Drouhard2017}. 
``Catalytic'' hackathons seek novel project ideas aimed at solving a tractable, well-defined challenge.
``Contributive'' hackathons seek to improve to an existing effort through focused work on discrete tasks, for example to make up for deficiencies in an ongoing project.
Finally, ``Communal'' hackathons place a strong focus on building a culture of practice and developing resources within an existing community, often defined by a specific domain of knowledge.

Our hack weeks extend the scientifically-focused, communal hackathon model into a space that includes a strong element of pedagogy and peer learning, forming the second axis of our events. 
Teaching within our events draws from a spectrum of approaches such as those traditionally employed in summer schools and software workshops. 
However, by embedding these approaches in a more open hackathon framework, we can explore unique ways of enhance the traditional summer school model.
For example, our hack weeks transfer knowledge across many levels of seniority, with participants routinely leading tutorials, and with numerous opportunities to `learn by doing' within a constructivist educational framework \cite{Bransford2000-lu}.
Additionally, the content of our tutorials tends to favor cutting-edge topics and experimental approaches over established knowledge. 
Finally, the social structure of hackathons enable us to prioritize participant diversity thereby maximizing the likelihood for the co-creation of learning during the event.

>>>>

Our hack weeks combine structured, tutorial-style instruction with informal education and peer learning opportunities occurring within projects and hacks. 
Our focus on both pedagogy and open project work aims to fill a space we feel is not currently addressed by existing models (Figure \ref{fig:hackspectrum}). 
The hackathon, a time-bounded, collaborative events that bring together participants around a shared challenge or learning objective \cite{Decker2015}, forms one primary axis of our events. 
Hackathons originated from the open-source software movement and have historically focused on software and technology development. 
In recent years hackathons have evolved into a model providing opportunities for intensive, interdisciplinary collaboration \cite{Groen2015-cj} and education \cite{Kienzler2015-zu,Lamers2014-xf} in the sciences. 
Core element of all hackathons include opportunities for networking, strengthening of social ties, and the building of community connections, both within and across disciplines.
Building on these core elements hackathons have been implemented in different ways depending on the overall purpose, mode of participation, style of work environment and participant motivation \cite{Drouhard2017}. 
``Catalytic'' hackathons seek novel project ideas aimed at solving a tractable, well-defined challenge.
``Contributive'' hackathons seek to improve to an existing effort through focused work on discrete tasks, for example to make up for deficiencies in an ongoing project.
Finally, ``Communal'' hackathons place a strong focus on building a culture of practice and developing resources within an existing community, often defined by a specific domain of knowledge.

Our hack weeks extend the scientifically-focused, communal hackathon model into a space that includes a strong element of pedagogy and peer learning, forming the second axis of our events. 
Teaching within our events draws from a spectrum of approaches such as those traditionally employed in summer schools and software workshops. 
By embedding these approaches in a more open hackathon framework, we can explore unique ways of enhancing the traditional summer school model.
For example, our hack weeks transfer knowledge across many levels of seniority, with participants routinely leading tutorials, and with numerous opportunities to `learn by doing' within a constructivist educational framework \cite{Bransford2000-lu}.
The content of our tutorials tends to favor cutting-edge topics and experimental approaches over established knowledge. 
Finally, the social structure of hackathons enable us to prioritize participant diversity thereby maximizing the likelihood for the co-creation of learning during the event.
